---
title: "ON Combined DEG troubleshooting"
author: "Jonathan Nelson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: bootstrapdata:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAABIAAAASCAYAAABWzo5XAAAAbElEQVR4Xs2RQQrAMAgEfZgf7W9LAguybljJpR3wEse5JOL3ZObDb4x1loDhHbBOFU6i2Ddnw2KNiXcdAXygJlwE8OFVBHDgKrLgSInN4WMe9iXiqIVsTMjH7z/GhNTEibOxQswcYIWYOR/zAjBJfiXh3jZ6AAAAAElFTkSuQmCC
    df_print: paged
    code_folding: hide
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r load_new_packages, echo=T, error=FALSE, message=FALSE, warning=FALSE}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("NatParksPalettes")) {install.packages("NatParksPalettes"); require("NatParksPalettes")}
if (!require("svglite")) {install.packages("svglite"); require("svglite")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("kableExtra")) {install.packages("kableExtra"); require("kableExtra")} # for color brewer

here()
```

```{r}
set.seed(2469)
```

# Helpful Objects
This section needs to be adjusted for each user. Either, have your own color scheme that matches your number of clusters, replicates, samples, age groups, and genotype groups -or- remove any specified color pallet in the graphing below. 
```{r echo=TRUE, warning=FALSE, error=FALSE, message=TRUE, fig.align = 'center'}
# Color palette for PIPseq that coordinates with Yellowstone for clusters
cluster_colors<- c("#0067A2", "#CC782B", "#5A8D66", "#509EA0", "#71A2B8", "#8A9BA9", "#B46DB3", "black")

# Color palette for replicates -- adapted from the Torres colors
replicate_colors <- c( "#7391BD" ,"#894846" ,"#E9988C" ,"#535260", "#B7A7A6" ,"#785838", "#C68D61" ,"#93995C")

# Color pallete for "disordered" sample_ID (in order of lexicon, 1, 10-16, 2-9)
sample_colors_d <- c("#C00000", "#FF8082", "#BFA85F", "#CCCA79", "#799E51", "#448A64", 
  "#568BB1", "#775790", "#FF0000", "#FFC000", "#F5EF1E", "#92D050", 
  "#00B050", "#0070C0", "#7030A0", "#D08B8C")

# Group color pallet
age_colors <- c("#BFA85F", "purple")

# genotype color pallet
genotype_colors <- c("orange", "blue")
```

# Increase Globals
```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

# Load Object

```{r, echo=TRUE, error=FALSE, fig.align= 'center', message=FALSE, warning=FALSE}
seu.obj<-readRDS(here("datasets", "3-Oligolineage_final.rds")) #Change to be your object

SO <- seu.obj

rm(seu.obj)

head(SO@meta.data)

SO

DimPlot(SO)

# Downsample because I'm on a computer with only 16GB of RAM (With Seurat v5)
# This code downsamples to 1000 cells for every active identity category (which in this case is Sample ID)

SO <- subset(SO, downsample = 1000)

rm(SO2)

```

# Testing Cluster DEGs

This seems to work fine after PrepSCTFindMArkers()

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(SO)

DimPlot(SO, group.by = "subclusters")

Idents(SO) <- SO@meta.data$subclusters

SO <- PrepSCTFindMarkers(SO)

df <- FindMarkers(SO, ident.1 = "OPC", logfc.threshold = 0.25, min.pct = 0.1)

VlnPlot(SO, "Pdgfra")

FeaturePlot(SO, "Pdgfra")

Idents(SO) <- SO@meta.data$subclusters

```

# Test Treatment DEGs

This doesn't work! Even though I know I just ran **PrepSCTFindMarkers()**

Error in FindMarkers.SCTAssay(object = data.use, latent.vars = latent.vars,  : 
  Object contains multiple models with unequal library sizes. Run `PrepSCTFindMarkers()` before running `FindMarkers()`.

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(SO)

# df2 <- FindMarkers(SO, ident.1 = "normal", group.by = "myelination", subset.ident = "OPC", logfc.threshold = 0.25, min.pct = 0.1)

```

# Solution

I think this problem stems from multiple issues. Mostly from new Seurat behaviors in V5 vs. V4 that we're not used to yet.

First, the **RNA assay layers** were never joined together, which I wonder if it messed up the initial SCT calculations. Regardless, there are multiple SCT assays in this object which I think is the heart of the problem. 

So my workaround is to: <br>
1) Set SO default assay to RNA and **join** all the layers <br>
2) Delete all previous SCT assays <br> 
3) Rerun SCT on the entire object <r>
<br>

I was worried that this would lose the UMAP embeddings, but it appears that those are still intact. 

```{r}

DefaultAssay(SO) <- "RNA"

SO <- JoinLayers(SO)

SO[["SCT"]] <- NULL

SO <- SCTransform(SO)

DimPlot(SO)

```


# Test Treatment DEGs

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(SO)

df2 <- FindMarkers(SO, ident.1 = "normal", group.by = "myelination", subset.ident = "OPC", logfc.threshold = 0.25, min.pct = 0.1)

VlnPlot(SO, "Srebf2", idents = "OPC", split.by = "myelination")

VlnPlot(SO, "Srebf2", idents = "OPC", split.by = "sample_ID", group.by = "myelination")

VlnPlot(SO, "Adamts9", idents = "OPC", split.by = "myelination")

VlnPlot(SO, "Adamts9", idents = "OPC", split.by = "sample_ID", group.by = "myelination")

```

# Session Info

```{r sessioninfo, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

sessionInfo()

```
