---
title: "ON Combined_psuedobulk anlysis "
author: "Katie Emberley"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: bootstrap
    df_print: paged
    code_folding: hide
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r load_new_packages, echo=T, error=FALSE, message=FALSE, warning=FALSE}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("NatParksPalettes")) {install.packages("NatParksPalettes"); require("NatParksPalettes")}
if (!require("svglite")) {install.packages("svglite"); require("svglite")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("kableExtra")) {install.packages("kableExtra"); require("kableExtra")} # for color brewer

here()
```

```{r}
set.seed(2469)
```

# Helpful Objects
This section needs to be adjusted for each user. Either, have your own color scheme that matches your number of clusters, replicates, samples, age groups, and genotype groups -or- remove any specified color pallet in the graphing below. 
```{r echo=TRUE, warning=FALSE, error=FALSE, message=TRUE, fig.align = 'center'}
# Color palette for PIPseq that coordinates with Yellowstone for clusters
cluster_colors<- c("#0067A2", "#CC782B", "#5A8D66", "#509EA0", "#71A2B8", "#8A9BA9", "#B46DB3", "black")

# Color palette for replicates -- adapted from the Torres colors
replicate_colors <- c( "#7391BD" ,"#894846" ,"#E9988C" ,"#535260", "#B7A7A6" ,"#785838", "#C68D61" ,"#93995C")

# Color pallete for "disordered" sample_ID (in order of lexicon, 1, 10-16, 2-9)
sample_colors_d <- c("#C00000", "#FF8082", "#BFA85F", "#CCCA79", "#799E51", "#448A64", 
  "#568BB1", "#775790", "#FF0000", "#FFC000", "#F5EF1E", "#92D050", 
  "#00B050", "#0070C0", "#7030A0", "#D08B8C")

# Group color pallet
age_colors <- c("#BFA85F", "purple")

# genotype color pallet
genotype_colors <- c("orange", "blue")
```

# Increase Globals
```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

# Load Object
```{r, echo=TRUE, error=FALSE, fig.align= 'center', message=FALSE, warning=FALSE}
seu.obj<-readRDS(here("datasets", "ALL_DCT.rds")) #Change to be your object

head(seu.obj@meta.data)

DimPlot(seu.obj, group.by = "Rep")

```

# Psuedobulk Object
## Group the cells and aggregate
Group by the cell just by sample_ID, not taking into account the independent clusters (we will use cluster information later and repeat this code).

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo <- AverageExpression(seu.obj, 
                           group.by = c("Rep"), # CHANGE the column with both condition and sample ID
                           assays = 'RNA', 
                           slot = "counts",
                           return.seurat = FALSE)

```

## Extract the counts
We will get a huge matrix with columns as the samples and rows as genes

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo <- pbo$RNA
pbo[1:6, 1:6]
colnames(pbo)[1:6]

pbo <- pbo*10000
pbo[1:6, 1:6]

```

## Add Metadata
This allows you to add whichever components of metadata you'd like to include to graph you PCA data by and whichever components you want to design the DESeq object from. 
```{r}
metadata <- seu.obj@meta.data %>%
  as.data.frame() %>%
  dplyr::select(Rep, Diet) # CHANGE These correspond to the column names of the metadata you want to extract 

dim(metadata)
head(metadata)

# Exclude duplicated rows 
metadata <- metadata[!duplicated(metadata), ]

dim(metadata)
metadata

# Rename rows to represent the sample IDs rather than the cell barcode IDs
rownames(metadata) <- metadata$Rep
head(metadata)

# This enusres the columns match up between pbo and the metadata. -- can probably skip this if your samples don't start with a number.
sample_names <- unique(seu.obj@meta.data$Rep) # CHANGE to be the column that you have for your samples
sample_names
colnames(pbo) <- sample_names
```

## Make DESeq Dataset
When making the DESeq dataset, make sure the countData have the exact same column names in the exacts same order as the rows of metadata. You get to choose your `design` for what the model will base it's sample matrix on. Select one or more of your metadata columns. 
```{r}
dds <- DESeqDataSetFromMatrix(countData = round(pbo), # Integers must be used for this matrix. 
                              colData = metadata,
                              design = ~ Diet) # CHANGE I am choosing to use sample ID rather than replicate here as we are trying to identify whether or not y20icKO batch differences will be an issue.

```

## Run PCA analyses
```{r, out.width = "100%"}
# Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)
```

### Plot PCA analyses
Change this section to customize based on which metadata you would like plotted. Note, this section includes several customizations that are user specific based on metadata and color preferences. 
```{r, out.width="100%"}

sample_names <- unique(rld$Rep)

# Plot PCA
sample<-BiocGenerics::plotPCA(rld, ntop = 1000, intgroup = "Rep") + 
  scale_color_manual(values = sample_colors_d)+ 
  geom_text_repel(aes(label = sample_names))+
  coord_fixed(ratio = 2.5) +  # Aspect ratio of 2 means the x-axis is twice as long as the y-axis
  theme_minimal()

plot(sample)

pcaData <- plotPCA(rld, ntop = 1000, intgroup = "Rep", returnData = TRUE)
pcaData

dds <- DESeq(dds)

res <- results(dds)
res

?plotPCA

PC1=pca$x[,1]
PC2=pca$x[,2]
PC3=pca$x[,3]

object <- rld

plotPCA <- function (object, intgroup = "Diet", ntop = 500, returnData = FALSE) 
{
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
        length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
        stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
        drop = FALSE])
    group <- if (length(intgroup) > 1) {
        factor(apply(intgroup.df, 1, paste, collapse = " : "))
    }
    else {
        colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4], group = group, 
        intgroup.df, name = colnames(object))
    if (returnData) {
        attr(d, "percentVar") <- percentVar[1:2]
        return(d)
    }
    ggplot(data = d, aes_string(x = "PC2", y = "PC3", color = "group")) + 
        geom_point(size = 3) + xlab(paste0("PC2: ", round(percentVar[1] * 
        100), "% variance")) + ylab(paste0("PC3: ", round(percentVar[2] * 
        100), "% variance")) + coord_fixed()
}

print(plotPCA(rld, intgroup=c("Diet")))

```

## fast_mnn

https://github.com/satijalab/seurat-wrappers/blob/master/docs/fast_mnn.md


```{r}

library(batchelor)
library(SeuratWrappers)

seu.obj <- NormalizeData(seu.obj)
seu.obj <- FindVariableFeatures(seu.obj)
seu.obj <- RunFastMNN(object.list = SplitObject(seu.obj, split.by = "Diet"))
seu.obj <- RunUMAP(seu.obj, reduction = "mnn", dims = 1:30)
seu.obj <- FindNeighbors(seu.obj, reduction = "mnn", dims = 1:30)
seu.obj <- FindClusters(seu.obj)
DimPlot(seu.obj, group.by = c("Diet", "ident", "seurat_annotations"), ncol = 3)

??RunFastMNN


```

# Genes Driving PCA
```{r}

rv = rowVars(assay(rld)) 
select = order(rv, decreasing=TRUE)[seq_len(min(500, length(rv)))]
pc = prcomp(t(assay(rld)[select,]))
loadings = as.data.frame(pc$rotation)
aload = abs(loadings)
sweep(aload, 2, colSums(aload), "/")
View(aload)

VlnPlot(seu.obj, features = "Coq10b", group.by = "Rep")
VlnPlot(seu.obj, features = "Spp2", group.by = "Rep")
VlnPlot(seu.obj, features = "Greb1", group.by = "Rep")
VlnPlot(seu.obj, features = "Pxdn", group.by = "Rep")

# Checking PCA

mat <- matrix(rexp(200, rate=.1), ncol=20)

project.pca <- prcomp(t(mat), scale=FALSE)

# Determine the proportion of variance of each component
# Proportion of variance equals (PC stdev^2) / (sum all PCs stdev^2)
project.pca.proportionvariances <- ((project.pca$sdev ^ 2) / (sum(project.pca$sdev ^ 2))) * 100

sum(project.pca.proportionvariances)

```

replicate<-DESeq2::plotPCA(rld, ntop = 1000, intgroup = "replicate") + 
  scale_color_manual(values = replicate_colors)+
  coord_fixed(ratio = 2.5) +  # Aspect ratio of 2 means the x-axis is twice as long as the y-axis
  theme_minimal()

age<-DESeq2::plotPCA(rld, ntop = 1000, intgroup = "age_group") + 
  scale_color_manual(values = age_colors)+
  coord_fixed(ratio = 2.5) +  # Aspect ratio of 2 means the x-axis is twice as long as the y-axis
  theme_minimal()


genotype<-DESeq2::plotPCA(rld, ntop = 1000, intgroup = "genotype") + 
  scale_color_manual(values = genotype_colors)+
  coord_fixed(ratio = 2.5) +  # Aspect ratio of 2 means the x-axis is twice as long as the y-axis
  theme_minimal()

batch<-DESeq2::plotPCA(rld, ntop = 1000, intgroup = "day_nuclei_isolated") + 
  scale_color_manual(values = c("pink", "darkgreen"))+
  coord_fixed(ratio = 2.5) +  # Aspect ratio of 2 means the x-axis is twice as long as the y-axis
  theme_minimal()


plot(sample)
plot(replicate)
plot(age)
plot(genotype)
plot(batch)


# Save the plots
folder<-here("4-Analysis", "PCA")

ggsave("PCA by Sample.svg", sample, path = folder)
ggsave("PCA by Replicate.svg", replicate, path = folder)
ggsave("PCA by Age.svg", age, path = folder)
ggsave("PCA by Genotype.svg", genotype, path = folder)
ggsave("PCA by Batch.svg", batch, path = folder)
```


# Other PCA Packages

## ggfortify

https://www.sthda.com/english/wiki/ggfortify-extension-to-ggplot2-to-handle-some-popular-packages-r-software-and-data-visualization

```{r}

library(ggfortify)

pca <- prcomp(pbo, scale. = TRUE)

autoplot(pca, loadings = TRUE, loadings.label = TRUE,
         data = pbo)

?prcomp

```

## FactoMiner

https://www.sthda.com/english/wiki/wiki.php?id_contents=7851

```{r}

library(FactoMineR)

PCA(pbo, scale.unit = TRUE, ncp = 5, graph = TRUE)

```



# Run PCA on a per cluster basis
# Psuedobulk Object
## Group the cells and aggregate
Group by the cell type and samples

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo <- AggregateExpression(seu.obj, 
                           group.by = c("fourth_label",     # CHANGE the column with the cell type information
                                        "sample_ID"), # CHANGE the column with both condition and sample ID
                           assays = 'RNA', 
                           slot = "counts",
                           return.seurat = FALSE)

```

## Extract the counts
We will get a huge matrix with columns as the cell_samples and rows as genes

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo <- pbo$RNA
pbo[1:16, 1:16]
colnames(pbo)[1:16]
```

### Data transformation
- Transpose the columns and rows
- Convert to data.frame

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo.t <- t(pbo)
pbo.t <- as.data.frame(pbo.t)
```

Check the structure of the data frame

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo.t[1:10, 1:10]
```

Split the rows (cell_samples) to remove the sample ID and only keep cell type information in the data frame
- Then we will get a vector containing the cell types, which will be used to split the data frame as a factor variable

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

# matching anything that's after the '_' and replace it with nothing (removing it)
splitRows <- gsub('_.*', '', rownames(pbo.t)) 
splitRows
```

Split data.frame using the splitRows vector
- It will return a list with each element corresponding with a cell type 
- Having data in such a list object makes it convenient to fetch matrices for any cell type in the data

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo.split <- split.data.frame(pbo.t,
                              f = factor(splitRows))
```

Check the matrix - using OPC as an example here

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

# rows as cell_sample ID and columns as genes
pbo.split$"OPC"[1:6, 1:6] # CHANGE 

```

Fix colnames and transpose it back to columns as the cell_samples and rows as genes

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

# use 'lapply' to remove the cell type and only leave the sample ID, and also re-transpose the matrix 
pbo.split.modified <- lapply(pbo.split, function(x){ 
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x)) # remove the cell type name from the row name
  t(x)
  
})
```

Check if you have performed the changes

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
pbo.split.modified$"OPC"[1:6, 1:6] # CHANGE 
pbo.split.modified$"Oligolineage"[1:6, 1:6] # CHANGE 
```

```{r, out.width = "100%"}
# List of clusters
clusters <- c("OPC", "Oligolineage", "Microglia", "Astrocyte", "Vascular")

# Loop over each cluster
for (cluster in clusters) {
  
  # Dynamically access the corresponding pbo.split.modified object (pbo.split.modified$<cluster>)
  pbo <- pbo.split.modified[[cluster]]
  
  # Ensure the columns of pbo match the sample names from your metadata
  sample_names <- unique(seu.obj@meta.data$sample_ID)
  colnames(pbo) <- sample_names

  # Create DESeq2 dataset
  dds <- DESeqDataSetFromMatrix(countData = round(pbo), 
                                colData = metadata,
                                design = ~ sample_ID)  # Change design formula if needed
  
  # Run PCA (Rlog transformation)
  rld <- rlog(dds, blind = TRUE)
  
  # Plot PCA by different metadata (you can customize the plots)
  sample_plot <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "sample_ID") + 
    scale_color_manual(values = sample_colors_d) + 
    geom_text_repel(aes(label = sample_names)) + 
    theme_minimal() +
    ggtitle(paste(cluster, "- sample_plot"))

  replicate_plot <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "replicate") + 
    scale_color_manual(values = replicate_colors) + 
    theme_minimal() +
    ggtitle(paste(cluster, "- replicate_plot"))

  age_plot <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "age_group") + 
    scale_color_manual(values = age_colors) + 
    theme_minimal() +
    ggtitle(paste(cluster, "- age_plot"))

  genotype_plot <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "genotype") + 
    scale_color_manual(values = genotype_colors) + 
    theme_minimal() +
    ggtitle(paste(cluster, "- genotype_plot"))

  batch_plot <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "day_nuclei_isolated") + 
    scale_color_manual(values = c("pink", "darkgreen")) + 
    theme_minimal() +
    ggtitle(paste(cluster, "- batch_plot"))

  # Print each plot to the console
  plot(sample_plot)
  plot(replicate_plot)
  plot(age_plot)
  plot(genotype_plot)
  plot(batch_plot)
  
  # Save the plots
  folder <- here("4-Analysis", "PCA")
  ggsave(paste0("PCA_", cluster, "_by_Sample.svg"), sample_plot, path = folder)
  ggsave(paste0("PCA_", cluster, "_by_Replicate.svg"), replicate_plot, path = folder)
  ggsave(paste0("PCA_", cluster, "_by_Age.svg"), age_plot, path = folder)
  ggsave(paste0("PCA_", cluster, "_by_Genotype.svg"), genotype_plot, path = folder)
  ggsave(paste0("PCA_", cluster, "_by_Batch.svg"), batch_plot, path = folder)

  # Optionally, print the current cluster being processed
  print(paste("Processing cluster:", cluster))
}
```

# Session Info

```{r sessioninfo, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
sessionInfo()
```
